---
title: "multipanel_code_along.Rmd"
author: "Peter Higgins"
date: "5/14/2020"
output: html_document
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

library(tidyverse)

remotes::install_github("thomasp85/patchwork")
library(patchwork)

remotes::install_github("wilkelab/ggtext")
library(ggtext)

knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      comment = NA)

```

## Preparing to Code-Along for Multipanel Plots

- make sure you have {tidyverse}, {patchwork}, and {ggtext} packages installed 

- library() all 3 packages

- Details at the {patchwork} packagedown site: <br>

https://patchwork.data-imaginist.com

- Take 60 seconds to install, library, and explore


## What Patchwork Can Do For You

- Start with Single (related) Plots

- Use a simple user interface to combine them

- Arrange at will in rows and columns

- Nesting included


## Preparing data for 4 dummy plots

```{r, echo=TRUE}
# Library calls
library(tidyverse)
library(ggtext)
library(patchwork)

# make dummy figures
d1 <- runif(500)
d2 <- rep(c("Treatment", "Control"), each=250)
d3 <- rbeta(500, shape1=100, shape2=3)
d4 <- d3 + rnorm(500, mean=0, sd=0.1)
plotData <- data.frame(d1, d2, d3, d4)
```


## Single Plots, Plot 1 through Plot 4

```{r, eval=FALSE}
p1 <- ggplot(data=plotData) + 
  geom_point(aes(x=d3, y=d4)) + labs(tag = "A") +
  theme(plot.background = 
          element_rect(color='black'))
p1

p2 <- ggplot(data=plotData) + geom_boxplot(aes(x=d2, y=d1, fill=d2))+
  theme(legend.position="none") + labs(tag = "B") +
  theme(plot.background = element_rect(color='black'))
p2

p3 <- ggplot(data=plotData) +
  geom_histogram(aes(x=d1, color=I("black"), fill=I("orchid"))) + labs(tag = "C") +
  theme(plot.background = element_rect(color='black'))
p3

p4 <- ggplot(data=plotData) +
  geom_histogram(aes(x=d3, color=I("black"), fill=I("goldenrod"))) + labs(tag = "D") +
  theme(plot.background = element_rect(color='black'))
p4
```

## Now Combine These Plots into a Multipanel Plot<br>with {patchwork}

### Use  |  to Combine Plots in a row!
```{r}

p1 <- ggplot(data=plotData) + 
  geom_point(aes(x=d3, y=d4)) + labs(tag = "A") +
  theme(plot.background = 
          element_rect(color='black'))

p2 <- ggplot(data=plotData) + geom_boxplot(aes(x=d2, y=d1, fill=d2))+
  theme(legend.position="none") + labs(tag = "B") +
  theme(plot.background = element_rect(color='black'))

p3 <- ggplot(data=plotData) +
  geom_histogram(aes(x=d1, color=I("black"), fill=I("orchid"))) + labs(tag = "C") +
  theme(plot.background = element_rect(color='black'))

p4 <- ggplot(data=plotData) +
  geom_histogram(aes(x=d3, color=I("black"), fill=I("goldenrod"))) + labs(tag = "D") +
  theme(plot.background = element_rect(color='black'))


p1 | p2 | p3 | p4
```


### Use  /  to Combine Plots in a Column!
```{r}
p1 <- ggplot(data=plotData) + 
  geom_point(aes(x=d3, y=d4)) + labs(tag = "A") +
  theme(plot.background = 
          element_rect(color='black'))

p2 <- ggplot(data=plotData) + geom_boxplot(aes(x=d2, y=d1, fill=d2))+
  theme(legend.position="none") + labs(tag = "B") +
  theme(plot.background = element_rect(color='black'))

p3 <- ggplot(data=plotData) +
  geom_histogram(aes(x=d1, color=I("black"), fill=I("orchid"))) + labs(tag = "C") +
  theme(plot.background = element_rect(color='black'))

p4 <- ggplot(data=plotData) +
  geom_histogram(aes(x=d3, color=I("black"), fill=I("goldenrod"))) + labs(tag = "D") +
  theme(plot.background = element_rect(color='black'))


p1 / p2 / p3 / p4 
```


### Use  +  to Combine Plots in a 2 x 2 Grid!
```{r}
p1 + p2 + p3 + p4
```


## Make More Complex Multipanel Plots with Nesting!
```{r}
p1 / (p2 | (p3/p4))
```


## Code-Along Break for Patchwork

- Look up the patchwork packagedown website, at
- https://patchwork.data-imaginist.com/

- Review the guides

- Try it out with plots p1 - p4

- Be creative with (), / , |, and plot_layout()

- Take 3 minutes to arrange some plots in the code chunk below

```{r}
p1 / ( p2 | p3 ) / p4
```

## This is Great, but How to Add a Combined Legend?


## The Side-Quest into {ggtext}:<br>This Turns Out to Be Part of the Solution!


## The {ggtext} package

- Packagedown site at https://wilkelab.org/ggtext/ 

- allows HTML tag formatting or markdown formatting of text within ggplots!

- color, italics, bold, images, boxes, superscripts

- Can be used in axis labels, legend text, annotations, tick labels

- uses element_markdown() or element_textbox() in place of element_text()

- multi-line labels with `<br>`

- can use inline CSS properties including color, font-size, and font-family


## Fun with {ggtext}: Replace Legends with Colored Text

```{r, echo = FALSE, fig.width=12}
mtcars %>%
  mutate(
    transmission = ifelse(am == 1, "automatic", "manual")
  ) %>%
  ggplot(aes(hp, mpg, color = transmission)) +
  geom_point(size = 2) +
  scale_color_manual(
    values = c(automatic = "#0072B2", manual = "#D55E00"),
    guide = "none"
  ) +
  labs(
    x = "Horse power",
    y = "Miles per gallon (MPG)",
    title = "<span style = 'font-size:14pt; font-family:Helvetica;'>Transmission type impacts fuel efficiency</span><br>MPG is higher for <span style = 'color:#0072B2;'>automatic</span>
than for <span style = 'color:#D55E00;'>manual</span> transmissions"
  ) +
  theme_bw() +
  theme(
    text = element_text(family = "Times"),
    plot.title.position = "plot",
    plot.title = element_markdown(size = 11, lineheight = 1.2)
  )
```


## Fun with {ggtext}: Add boxes to axes, titles
```{r, echo = FALSE, fig.width=12}

p <- iris %>%
  filter(Species == 'versicolor') %>% 
  ggplot() +
  aes(x= Sepal.Width, y = Sepal.Length) +
  geom_point()+
  theme_minimal() +
  geom_smooth(method = 'lm') +
  theme(plot.background = element_rect(color='black')) 


p + labs(
  title = "<b>The Unmitigated Awesomeness of Statins</b>",
  caption = "<b>Figure 1.</b>
    <span style = 'font-size:10pt'>This is an Amazing Plot. *Look at this
    preliminary data*, **It is great.**
    <span style = 'color:red;'>This work will advance the field,</span>
    You should *definitely* fund this grant.</span>",
  x = "Exposure to statin medications (years * mg/d / 50)",
  y = "Survival (years)<br><span style = 'font-size:8pt'>A measure of
    the effect of exposure.</span>"
) +
  theme(
    plot.title.position = "plot",
    plot.title = element_textbox_simple(
      size = 15,
      lineheight = 1,
      padding = margin(4, 5.5, 4, 5.5), #padding inside the box, (trouble)
      margin = margin(0, 0, 7, 0), #margins outside the box, t,r,b,l
      linetype = 1,
      r = grid::unit(0, "pt"),
      fill = 'gray90'
    ),
    plot.caption.position = "plot",
    plot.caption = element_textbox_simple(
      size = 11,
      lineheight = 1.1,
      padding = margin(5.5, 5.5, 5.5, 5.5),
      margin = margin(0, 0, 0, 0),
      r = grid::unit(0, "pt"),
      linetype = 1
    ),
    axis.title.x = element_textbox_simple(
      width = NULL,
      padding = margin(4, 4, 4, 4),
      margin = margin(4, 0, 4, 0),
      linetype = 1,
      r = grid::unit(8, "pt"),
      fill = "azure1"
    ),
    axis.title.y = element_textbox_simple(
      hjust = 0,
      orientation = "left-rotated",
      minwidth = unit(1, "in"),
      maxwidth = unit(2, "in"),
      padding = margin(4, 4, 2, 4),
      margin = margin(0, 0, 2, 0),
      fill = "lightsteelblue1"
    )
  )
```


## Returning to The Quest to Add a Combined Legend


### Failed Attempts

- With a TextGrob - bad

- with a {ggtext} caption - not great

- with a {ggtext} textbox_grob - closer
  - Odd spacing
  - Turns out each row set to equal height
  
- with a {ggtext} plot_annotation - very close!

## Code and Image - {ggtext} plot_annotation 

```{r}
fig_legend <- plot_annotation(
  caption = "**Figure 1.**  Testing Control vs. Treatment.   A. Scatterplot. 
  B. The outcomes in the control arm were significantly better than 
  the Treatment Arm. C. Histogram. D. Another Histogram.",
  theme = theme(
    plot.caption = element_textbox_simple(
      size = 11,
      box.colour = "black",
      linetype = 1,
      padding = unit(c(3, 3, 3, 3), "pt"),
      r = unit(0, "pt")  )  ) )
p1 + {
  p2 + {
    p3 +
      p4 +
      plot_layout(ncol=1)
  }
} + fig_legend +
  plot_layout(ncol=1)
```




## Now Add the Legend - with a {ggtext} plot_annotation 

- Very close, but funky margins (narrower plot_annotation)

- turns out that this inherits margins from plot margins above - now applied to the caption

- can assign negative left and right margins (5.5 pt) to counteract the default plot margins.


## Now Add the Legend - Attempt 5 - Solution!

- With negative left and right margins (5.5 pt) 

```{r}
fig_legend <- plot_annotation(
  caption = "**Figure 1.**  Testing Control vs. Treatment.   A. Scatterplot. 
  B. The outcomes in the control arm were significantly better than 
  the Treatment Arm. C. Histogram. D. Another Histogram.",
  theme = theme(
    plot.caption = element_textbox_simple(
      size = 11,
      box.colour = "black",
      linetype = 1,
      padding = unit(c(3, 3, 3, 3), "pt"),
      margin = unit(c(0, -5.5, 0, -5.5), "pt"), 
      r = unit(0, "pt")  ) ) )

p1 + {
  p2 + {
    p3 + p4 + plot_layout(ncol=1)}} + 
  fig_legend +
  plot_layout(ncol=1)
```



## Lessons Learned

- The #rstats community is awesome

- Use {reprex} to ask questions on RStudio Community and Stack Overflow

- Lots of people are willing to help if you can come up with a *minimal* reproducible example and<br>can state your question (and your end goal) clearly

# Links to RStudio Community and StackOverflow Discussions

<br><br>
[RStudio Community](https://community.rstudio.com/t/plot-in-a-box-with-long-text-legend-in-a-box/51962)
<br>
https://community.rstudio.com/t/plot-in-a-box-with-long-text-legend-in-a-box/51962
<br>

[Stack Overflow](https://stackoverflow.com/questions/60142452/multipanel-plots-with-spanning-figure-legend-in-r-with-ggtext-and-gridtext-in-r)
<br>
https://stackoverflow.com/questions/60142452/multipanel-plots-with-spanning-figure-legend-in-r-with-ggtext-and-gridtext-in-r


## Other Cool Stuff I Used that Is Worth Checking Out

<br>
First time using {xaringan} for a presentation
<br>
https://github.com/yihui/xaringan
<br>

The {xaringanthemer} package for the maize and blue color scheme and fonts
<br>
https://pkg.garrickadenbuie.com/xaringanthemer/
<br>

The {countdown} package for the timer
<br>
https://github.com/gadenbuie/countdown
<br>

Side-by-Side Code and Plots in {xaringan}<br>
https://www.garrickadenbuie.com/blog/decouple-code-and-output-in-xaringan-slides/
<br>

Code Highlighting in {xaringan}<br>
https://www.garrickadenbuie.com/blog/highlight-lines-without-breaking-the-code-in-xaringan/
<br>

{reprex} for minimal reproducible examples to ask questions <br>
https://reprex.tidyverse.org/ 

